TacticToe: Learning the Game of Tactics

### 2018-01-29 tactictoe_jv.tex
minor

\section{Preliminaries: Search Tree}\label{sec:prelim}

@@

Starting from the bottom parent node, we can construct children of a goal of 
this node by choosing a goal $g_i^0$ on this node and choosing tactics 
$t_1,t_2,\ldots$ to 
apply to this goal. Here the result produces new nodes each containing a list 
of goals to be proven. In each node of the proof tree, a goal can either be 
solved, open or pending (see Definition~\ref{def:solved}).

\begin{definition}\label{def:solved}(Solved goal, solved nodes, open goal, 
pending goal)\\
Let $\mathfrak{T}$ be a search tree. The set of solved nodes 
$\mathbb{A}^\infty$ and 
the set of solved goals $\mathbb{G}^\infty$ are defined inductively by:

\begin{align*}
\mathbb{A}^{0} &=_{def} 
\lbrace a \in \mathbb{A}\ |\ G(a) = \emptyset \rbrace \\ 
\mathbb{G}^{0} &=_{def} \lbrace g \in \mathbb{G}\ |\ 
\exists a \in A(g).\ a \in \mathbb{A}^{0} \rbrace\\
\mathbb{A}^{n+1} &=_{def} \lbrace a \in \mathbb{A}\ |\ 
\forall g \in G(a).\ g \in \mathbb{G}^{n} \rbrace\\
\mathbb{G}^{n+1} &=_{def} \lbrace g \in \mathbb{G}\ |\ 
\exists a \in A(g).\ a \in \mathbb{A}^{n+1} \rbrace \\
\mathbb{A}^\infty &=_{def} \bigcup_{i \in \mathbb{N}} \mathbb{A}^n \ \ \ \ \ 
\mathbb{G}^\infty =_{def} \bigcup_{i \in \mathbb{N}} \mathbb{G}^n\\
\end{align*}

@@

The set of pending goals is the remaining set of goals in the tree. They are 
goal that were produced by tactics that were not explored yet. This set can 
also be also defined by:
\[\lbrace g \in G\ |\  T(g) = \emptyset \rbrace\]
\end{definition}

@@

%In this section, we first give a static description of the proof tree and its 
%component. 
We first explain how to compute the policy and evaluation functions that 
influence the behaviour of the MCTS algorithm.
Then we describe the five steps of the MCTS algorithm and their effect on 
the search tree (described in Section~\ref{sec:prelim}): 
initialization, node selection, node extension, backpropagation and resolution. 
A characteristic of the MCTS algorithm is that it offers a good trade-off 
between exploitation and exploration. This means that the algorithm searches
deeper more promises branches and leaves enough time for the exploration 
of less likely alternatives.

\subsection{Prior tactic policy}\label{sec:policy}
Distances between goals produced by the tactic predictor are hard 
to translate into a prior tactic policy, so we rely only on the order of 
tactics to estimate the probability with which each tactic should be tried.
Let $t_0,\ldots,t_500$ be the list of preselected tactics ordered by their 
distance to a goal $g$. Let $c_{policy}$ be a constant heuristic that estimates 
how likely a tactic $t_i$ is to prove the goal compared to the rest of the 
tactics $t_{i+1},\ldots,t_{500}$. We make the strong assumption that $c_policy$ 
is constant across the search. We can now calculate the prior policy by:

\[PriorPolicy(t_i) = PP(i) = (1 - c_{policy})^{i} * c_{policy}\]

The first term of the multiplication is the probability that none of the better 
ranked tactic is selected.

In fact, these probabilities are never used directly and only serves as a 
guidance to our deterministic node selection algorithm over many simulations.

Since the support of A induces a bijection between tactics of the tree 
search and the node they 
produced, any function define on tactics of the tree search can be map to a 
function defined on nodes. 
Therefore, we define $PriorPolicy(a)$ as the $PriorPolicy(t)$ where $t$ is the 
tactic that produced $a$.

@@

We now concentrate on defining a reasonable evaluation function that also 
influences node selection in MCTS. Essentially, we estimate how 
interesting is a list of goals produced by a tactic. This way, we can determine 
if we should continue searching in this direction. Theoretically, an 
interesting list of goal is one 
that has the shortest proof, and we could measure the likelihood of being 
provable in a number of steps from previous proof search. However, this 
approach produces a lot of negatives\todo{TG: cite Hard Negative mining (cite 
Deep math)}. So to reduce this number, we limit 
ourselves to tactics tested during orthogonalization. We declare a list 
of goals l to be positive, noted $l \in Positive$, if it is has been 
produced by the winner of an orthogonalization competition. The set of the $r$ 
goals closest to a list of goals $l$ is noted $Neighbour(r,l)$. We can 
define the set of positives within a 
radius $r$ of $l$ to be $Positive(r,l) = \lbrace m \in Neighbour(r,l)\ |\ m \in 
Positive \rbrace$.

@@

Through node extension steps the search tree grows and the number of paths to 
explore increases. To decide which node to extend next, we proceed through the 
node selection algorithm. Starting from the root node, we repeatedly chose the 
children with the highest value.

The value of a $a_i$ with respect to a search tree is determined by:
\[Value(a_i) = CurrentEvaluation(a_i) + c_{exploration} * Exploration(a_i)\]

The exploration term is determined by the prior policy and the current policy 
where as the exploitation is exactly the current evaluation build from prior of 
evaluations of descendants of $a_i$.
\[Exploration(a_i) = \frac{PriorPolicy(a_i)}{CurrentPolicy(a_i)}\]

\[CurrentPolicy(a_i) = \frac{1 + Visit(a_i))}{\sqrt{Visit(p))}}\]
The policy is approximatively the  percentage of time a node was visited.The 
square root skews this probability to favor even more exploration of nodes with 
few visits.

The exploitation term is the average of the evaluation of 
all nodes descendant of $a_i$ (itself included) in the current search tree.
This evaluation also include failures of the selection process.

 It 
is also called the current evaluation of $a_i$.   
\[CurrentEvaluation(a_i) = 
  \sum_{b \in Desc(a_i)} \frac{PriorEvaluation(b)} {(card Desc(a_i) + card 
  (Failure(a_i))}\]

where $Failure(a_i)$ is the number of run that went through $a_i$ that ended in 
a failure.

To allow the algorithm to select nodes that 
are not leafs, we also assign a value to the possibility of stopping during the 
descent of the tree. This value has an exploitation equal to 0 since it was 
never explored and is given a prior policy equal to $PP(n)$ where $n$ is the 
number of 
tactics already tried on the open goal if there is one and $PP(0)$ if there is 
none.

\subsection{Node extension}
Once a node is selected, we applied the best not-yet-applied tactics according 
to the prediction order for the open goal of this node.
If no open goals exists we apply it to the first pending goal (according to 
some arbitrary order). If the node is a solved node or (a descendant of a 
solved node) then 
If the tactic is productive then the we create a new 
node and calculate and save its prior evaluation. Otherwise we evaluate the 
output of the tactic to 0 and backpropagate the failure to the ancestries.

Maybe should move this definition in static description.

To prevent tactics from looping, we interrupt tactics after
0.01 to 0.1 seconds. The optimal tactic timeout varies in function of the 
search time limit.

\begin{definition} (Productive tactic)
During a proof search, after a tactic $t$ is applied to a node with active goal 
$g$, $t$ is called 
productive if and only if all the following conditions are satisified:
\begin{itemize}
\item It does not fail. We note $l$ the list of goals returned by the tactic.
\item It does not loop. $l$ do not contain any goals that appears in its 
ancestor nodes.
\item It is not a parallel step. $l$ is not equal to or a superset of a list of 
goals produced from the same node with the same active goal.
\end{itemize}

The third point of the definition is a partial attempt to prevent confluent 
branches. The general 
case where two branches becomes parallel after $n$ steps is handle by a
tactic cache which memorizes tactic effects on goals. Similarly there is also a 
prediction cache, which memorize the tactic predictions for each goal. Those 
two caches allow a fast exploration of confluent branches but they remain 
separated in our search tree.

\subsection{Backpropagation}
Number of failed selection and 
the number of visits are backpropagated to the ancestors

which update the current evaluation in each node.

\subsection{Resolution}
The search ends when the algorithm reaches one of these 3 conditions:
\begin{itemize}
\item It finds a proof (i.e. the root node is solved). In this case, 
the search returns a minimized and 
prettified proof script (see Section~\ref{sec:proofdisplay}).
\item It saturates. There is no tactics to be applied to any open goal.
\item It exceeds a given time limit.
\end{itemize}

The second case happens very rarely (less than 10 times in the full scale 
experiment). So our experiments report only the number of proofs found relative 
to the number of attempted problems.
\end{definition}

>>>>

\section{Preliminaries: Search Tree}\label{sec:prelim}
\todoi{Print words correctly in math format by avoiding math format}

@@

\begin{definition}\label{def:desc}(Children, Descendant)\\
We now define what are the children and the descendants of a node $a$ in a 
search tree $\mathfrak{T}$. The children of a relative to a goal $g$ in $G(a)$ 
is the set of nodes produced by tactics applied to $g$:
  \[Children(a,g) = \lbrace A(g,t)\in \mathbb{A}\ |\ t \in T(g) \rbrace \]

From that, we define the children of a node $a$ by:
  \[Children(a) = \bigcup_{g \in G(a)} Children(a,g) \]

And descendants of $a$ by transitive closure:
\begin{align*}
Descendant^{0}(a) &=_{def} \lbrace a \rbrace \\ 
Descendant^{n+1}(a) &=_{def} \bigcup_{a' \in Descendant^{n}(a)} 
Children(a') \\
Descendant(a) &=_{def} \bigcup_{n \in \mathbb{N}} Descendant^n(a)\\
\end{align*}

Starting from the bottom parent node, we can construct children of a goal of 
this node by choosing a goal $g_i^0$ on this node and choosing tactics 
$t_1,t_2,\ldots$ to 
apply to this goal. Here the result produces new nodes each containing a list 
of goals to be proven. In each node of the proof tree, a goal can either be 
solved, open or pending (see Definition~\ref{def:solved}). 
\end{definition}

\begin{definition}\label{def:solved}(Solved goal, solved nodes, open goal, 
pending goal , active node)\\
Let $\mathfrak{T}$ be a search tree. The set of solved nodes 
$\mathbb{A}^\infty$ and 
the set of solved goals $\mathbb{G}^\infty$ are defined inductively by:

\begin{align*}
\mathbb{A}^{0} &=_{def} 
\lbrace a \in \mathbb{A}\ |\ G(a) = \emptyset \rbrace \\ 
\mathbb{G}^{0} &=_{def} \lbrace g \in \mathbb{G}\ |\ 
\exists a \in A(g).\ a \in \mathbb{A}^{0} \rbrace\\
\mathbb{A}^{n+1} &=_{def} \lbrace a \in \mathbb{A}\ |\ 
\forall g \in G(a).\ g \in \mathbb{G}^{n} \rbrace\\
\mathbb{G}^{n+1} &=_{def} \lbrace g \in \mathbb{G}\ |\ 
\exists a \in A(g).\ a \in \mathbb{A}^{n+1} \rbrace \\
\mathbb{A}^\infty &=_{def} \bigcup_{n \in \mathbb{N}} \mathbb{A}^n \ \ \ \ \ 
\mathbb{G}^\infty =_{def} \bigcup_{n \in \mathbb{N}} \mathbb{G}^n\\
\end{align*}

@@

The set of pending goals is the remaining set of goals in the tree. They are 
goal that were produced by tactics that were not explored yet. This set can 
also be also defined by:
\[\lbrace g \in G\ |\  T(g) = \emptyset \rbrace\]

To facilitate the description  of our algorithms we make a slight adjustement 
to the definition of an open goal. If a node $a$ does not contain any open 
goal, the first pending goal (according 
to some preset order) of $a$ is also considered to be an open goal. In this 
way, each unsolved node contains a unique open goal.
\end{definition}

\todo{make definition in preliminaries of active node and descendant?}

We can also judge the productivity of a tactic by its contribution to the 
search 
tree.

\begin{definition} (Productive tactic) 
The application of a tactic on an open goal is called productive if and only if 
all the following conditions are satisified:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we 
interrupt tactics after a short amount of time (0.05 seconds). 
\item It does not loop. Its output does not contain any goals that appears in 
its ancestor nodes.
\item It is not a parallel step. Its output is not equal or a superset 
of a list of goals produced from the same open goal.
\end{itemize}

The third point of the definition is a partial attempt at preventing confluent 
branches. The general case where two branches join after $n$ steps is handled 
by a tactic cache which memorizes tactic effects on goals. This
cache allows faster re-exploration of confluent branches which remain 
separated in the search tree.

\todo{make definition in preliminaries of active node and descendant?}

@@

%In this section, we first give a static description of the proof tree and its 
%component. 
We first explain how to compute the policy and evaluation functions that 
influence the behavior of the MCTS algorithm.
We then describe initialization and resolution of the search as well as the 
different part of a step in the search loop: node selection, node 
extension, backpropagation.
A characteristic of the MCTS algorithm is that it offers a good trade-off 
between exploitation and exploration. This means that the algorithm searches
deeper more promises branches and leaves enough time for the exploration 
of less likely alternatives.

\subsection{Prior tactic policy}\label{sec:policy}

\todoi{mention productive tactics}
Distances between goals produced by the tactic predictor are hard 
to translate into a prior tactic policy, so we rely solely on the order of 
tactics to estimate the probability with which each tactic should be tried.
Let $t_0,\ldots,t_n$ be the list of preselected tactics ordered by their 
distance to a goal $g$. Let $c_{policy}$ be a constant heuristic that estimates 
how likely a tactic $t_i$ is to prove the goal compared to the rest of the 
tactics $t_{i+1},\ldots,t_{500}$. We make the strong assumption that $c_policy$ 
is constant across the search. We can now calculate the prior policy by:

\[PriorPolicy(a_i) = PriorPolicy(t_i) = PriorPolicy(i) = (1 - c_{policy})^{i} * 
c_{policy}\]

Widening policy for a node $a$.

\[WideningPolicy(a) = \]

The first term of the multiplication is the probability that none of the better 
ranked tactic is selected. In fact, these probabilities are never used directly 
and only serves as a guidance to our deterministic node selection algorithm 
over many MCTS step.

@@

We now concentrate on defining a reasonable evaluation function that also 
influences node selection in MCTS. Essentially, we estimate how 
interesting is a list of goals produced by a tactic. This way, we can determine 
if we should continue searching in this direction. Theoretically, an 
interesting list of goal is one 
that has the shortest proof, and we could measure the likelihood of being 
provable in a number of steps from previous proof search. However, this 
approach produces a lot of negatives\todo{CK: cite hard negative mining 
or/and deep math}. So to reduce this number, we limit 
ourselves to tactics tested during orthogonalization. We declare a list 
of goals l to be positive, noted $l \in Positive$, if it is has been 
produced by the winner of an orthogonalization competition. The set of the $r$ 
goals closest to a list of goals $l$ is noted $Neighbour(r,l)$. We can 
define the set of positives within a 
radius $r$ of $l$ to be $Positive(r,l) = \lbrace m \in Neighbour(r,l)\ |\ m \in 
Positive \rbrace$.

@@

Through node extension steps the search tree grows and the number of paths to 
explore increases. To decide which node to extend next, we proceed through the 
node selection algorithm. Starting from the root node, we repeatedly chose the 
children relative to the open goal with the highest value.

The value of a $a_i$ with respect to a search tree is determined by:
\[Value(a_i) = CurEvaluation(a_i) + c_{exploration} * Exploration(a_i)\]

The exploration term is determined by the prior policy and the current policy 
where as the exploitation is exactly the current evaluation build from prior of 
evaluations of descendants of $a_i$.
\[Exploration(a_i) = \frac{PriorPolicy(a_i)}{CurPolicy(a_i)}\]

\[CurPolicy(a_i) = \frac{1 + Visit(a_i))}{\sqrt{Visit(p))}}\]
The policy is approximatively the  percentage of time a node was visited.The 
square root skews this probability to favor even more exploration of nodes with 
few visits.

The exploitation term $CurEvaluation(a_i)$ is the average of the evaluation of 
all MCTS step that went through $a_i$ (itself included) in the current search 
tree.

\[CurEvaluation(a_i) = 
  \sum_{a' \in Descendant(a_i)} \frac{PriorEvaluation(a')} {card\ 
  Descendant(a_i) 
  + card\ Failure(a_i)}\]

where $Failure(a_i)$ is the number of MCTS steps that went through $a_i$ which 
ended in a failure.

To allow the algorithm to select nodes that 
are not leafs, we also assign a value to the possibility of stopping during the 
descent of the tree. This value has an exploitation equal to 0 since it was 
never explored and is given a prior policy equal to $PriorPolicy(n)$ where $n$ 
is the 
number of 
tactics already tried on the open goal if there is one and $PriorPolicy(0)$ if 
there is 
none.

\subsection{Node extension}
Once a node is selected, we first check if it is an active part of the search 
tree (not a solved node .

If the selected node is not active, then the result of node extension is a 
failure.
Otherwise, we apply the best untested tactic on the open goal $g$ of this node  
according to the prediction order for $g$.

we apply the next tactic in the prediction order for 
the open goal of this node.

The result of the extension step is either a failure or a list of goals $l$.
From $l$, we can assess if the tactic was productive. 
If not, we also return a failure.
Otherwise, the list of goals $l$ is used to create a new node in the search 
tree. Give precise definition.

order of predicted tactics.

During the creation of node, we predict tactics for the open goals

\subsection{Backpropagation}

If node extension resulted in a failure, we increment the failure field of the 
selected node and its parents. If node extension was successful, we 
backpropagate the evaluation of the new nodes to their parents.
In both cases, we increment the number of visits of all nodes traversed.
This changes induce an update on the current evaluation and current policy for 
each node. And the process loops back to the node selection step.

\subsection{Resolution}
The search ends when the algorithm reaches one of these 3 conditions:
\begin{itemize}
\item It finds a proof (i.e. the root node is solved). In this case, 
the search returns a minimized and 
prettified proof script (see Section~\ref{sec:proofdisplay}).
\item It saturates. There is no tactics to be applied to any open goal.
This happens less than 10 times in the full-scale experiment.
\item It exceeds a given time limit set by default to 60 seconds.
\end{itemize}

\end{definition}
